# Set up

```{r setup-echo-setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

The command-line chapters of this workshop (`gprofiler2`, `clusterProfiler` and `enrichplot`, Novel species FEA, `FGSEA`, and `STRINGdb`) use a number of R packages. On the day, these are already installed on the workshop-provided computers/VMs, so no action is needed. This page is for anyone who wants to run the code on their **own computer**, either during the workshop or afterwards.

This workshop was built and tested with **R 4.6.1** and **Bioconductor 3.23**. Any recent R release (4.4 or later) with a matching Bioconductor release should work; see the [Bioconductor release/R version table](https://bioconductor.org/about/release-announcements/) if you are installing on an older R version.

<p>&nbsp;</p>  <!-- insert blank line -->

## CRAN packages

Most of these come from the [tidyverse](https://www.tidyverse.org/) plus a handful of enrichment-specific packages:

```{r install-cran, eval = FALSE}
install.packages(c(
  "tidyverse", # includes dplyr, ggplot2, readr, tidyr, and friends
  "data.table",
  "gprofiler2",
  "ggupset",
  "ggtangle",
  "ggridges", # needed by enrichplot::ridgeplot()
  "msigdbr",
  "ontologyIndex",
  "WebGestaltR",
  "DT",
  "ggrepel"
))
```

**Note on `WebGestaltR`:** it includes Rust source code that gets compiled during installation, so a working Rust toolchain (including `cargo`) is required beforehand. If installation fails with a "cargo command is not available" or similar error, install/update Rust via [rustup](https://www.rust-lang.org/tools/install) rather than an older distro-packaged `cargo`, then retry `install.packages("WebGestaltR")`.

<p>&nbsp;</p>  <!-- insert blank line -->

## Bioconductor packages

Bioconductor packages are installed via `BiocManager` rather than `install.packages()`:

```{r install-bioc, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c(
  "clusterProfiler",
  "enrichplot",
  "biomaRt",
  "org.Hs.eg.db",
  "fgsea",
  "BiocParallel",
  "STRINGdb",
  "ReactomePA",
  "AnnotationDbi",
  "graphite"
))
```

`org.Hs.eg.db` is the human genome-wide annotation package; it matches the human [Pezzini et al. 2017](https://link.springer.com/article/10.1007/s10571-016-0403-y) dataset used throughout the workshop.

<p>&nbsp;</p>  <!-- insert blank line -->

## Optional: GitHub-only package

The `clusterProfiler`/`enrichplot` chapter mentions `multienrichjam`, a package that converts ORA results from other tools into `enrichplot`-compatible objects. It is **not required** for this workshop, has a large dependency footprint, and is not installed on the workshop VMs. If you'd like to try it on your own computer:

```{r install-multienrichjam, eval = FALSE}
# install.packages("remotes")
# remotes::install_github("jmw86069/multienrichjam", dependencies = TRUE)
```

<p>&nbsp;</p>  <!-- insert blank line -->



## Raw Data

```{r }
# where are you? what folder are you working in

getwd()

## Download and untar the data
options(timeout = 3600)
download.file(
  "https://bioinformatics.erc.monash.edu/home/hval0004/FEA_Workshop/data.tar", "data.tar"
)

untar("data.tar")

## Download workshop.R script. This file contains all the example code for this workshop.
download.file(
  "https://monashbioinformaticsplatform.github.io/Functional_Enrichment_Web_and_Command_Line/workshop.R", "workshop.R"
)

# list downlowded files
list.files()
list.files("data")
```