# Overview

## Functional Analysis of -Omics Data

### General Information

The workshop covers the bioinformatics concepts and tools for interpreting a gene list using Gene Ontology and pathway information. The workshop focuses on the principles and concepts required for analysing and conducting functional and pathway analyses on a gene list from any organism, although the focus will be on human and model eukaryotic organisms.

### Course Objectives

Participants will gain practical experience and skills to be able to:

-   Understand basic concepts of functional enrichment analysis;
-   Interpret enrichment analysis results;
-   Gain systems perspective of gene functions;
-   Obtain more information about a gene list;
-   Discover which pathways are enriched in a gene list (and use this information for hypothesis generation);
-   Predict gene function and extend a gene list;
-   Follow the workflow after the workshop to conduct their own analyses.

### Target Audience

This workshop is intended for biologists working with ‘-Omics data’ (e.g., RNA-seq, protein expression), who are interested in interpreting large gene/protein lists resulting from their experiments.

### Setup Requirements

The workshop will be held **in person**. Participants are encouraged to bring a **laptop**; however, **no software installation is necessary**.  

Workshop materials and relevant links can be found on the [**Resources**](https://monashbioinformaticsplatform.github.io/Functional_Enrichment_Web_and_Command_Line/resources.html) page.


BYO coffee :)

### Schedule - Day 1

```{r, echo=FALSE, message=FALSE}
library(knitr)
suppressWarnings(library(kableExtra))
library(downlit)

library(dplyr)
library(lubridate)

start_time <- ymd_hm("2026-08-12 09:00") # anchors the day

schedule <- tibble::tibble(
  Instructor = c(
    # LP (10 min)
    "LP",
    # HK (3 items before break)
    rep("HK", 3),
    # Morning break (10 min)
    "",
    # GI - GSEA (web) + fgsea
    "GI",
    # GI - STRING (web) + STRINGdb
    "GI",
    # Lunch
    "",
    # HK - gProfiler (web) + gprofiler2
    "HK",
    # SA - Reactome (web) + ReactomePA
    "SA",
    # Wrap-up / Q&A / preview Day 2
    ""
  ),
  Activity = c(
    "Welcome and housekeeping",
    "Introduction to functional enrichment analysis",
    "Data acquisition & filtering gene list",
    "Interactive Calculator",
    "Morning break",
    "GSEA (web) + fgsea",
    "STRING (web) + STRINGdb",
    "Lunch break",
    "g:Profiler (web) + gprofiler2",
    "Reactome (web) + ReactomePA",
    "Wrap-up / Q&A / preview Day 2"
  ),
  `Time (mins)` = c(
    10, # LP
    35, 10, 15, # HK before break
    10, # Morning break
    60, # GI: GSEA (web) + fgsea
    50, # GI: STRING (web) + STRINGdb
    40, # Lunch break
    60, # HK: g:Profiler (web) + gprofiler2
    60, # SA: Reactome (web) + ReactomePA
    10 # Wrap-up / Q&A / preview Day 2
  )
) %>%
  mutate(
    Start = start_time + minutes(c(0, head(cumsum(`Time (mins)`), -1))),
    End   = start_time + minutes(cumsum(`Time (mins)`)),
    Start = format(Start, "%I:%M %p"),
    End   = format(End, "%I:%M %p")
  )

# Sanity check: full 360-min day (9:00–15:00)
stopifnot(sum(schedule$`Time (mins)`) == 360)

# Render the table
kable(schedule, "html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(1, width = "10em", bold = TRUE) %>%
  column_spec(3, width = "8em")
```

### Schedule - Day 2

```{r, echo=FALSE, message=FALSE}
start_time_day2 <- ymd_hm("2026-08-13 09:00") # anchors day 2

schedule_day2 <- tibble::tibble(
  Instructor = c(
    "", # Recap of Day 1 & Day 2 overview
    "SA", # clusterProfiler and enrichplot
    "", # Morning break
    "HK", # Novel species FEA
    "" # Wrap-up
  ),
  Activity = c(
    "Recap of Day 1 & Day 2 overview",
    "clusterProfiler and enrichplot",
    "Morning break",
    "Novel species FEA",
    "Wrap-up: Uncertainties, Reporting & Resources (self-paced) + Q&A"
  ),
  `Time (mins)` = c(
    10, # Recap
    65, # SA: clusterProfiler and enrichplot
    15, # Morning break
    80, # HK: Novel species FEA
    10 # Wrap-up
  )
) %>%
  mutate(
    Start = start_time_day2 + minutes(c(0, head(cumsum(`Time (mins)`), -1))),
    End   = start_time_day2 + minutes(cumsum(`Time (mins)`)),
    Start = format(Start, "%I:%M %p"),
    End   = format(End, "%I:%M %p")
  )

# Sanity check: full 180-min morning (9:00-12:00)
stopifnot(sum(schedule_day2$`Time (mins)`) == 180)

# Render the table
kable(schedule_day2, "html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(1, width = "10em", bold = TRUE) %>%
  column_spec(3, width = "8em")
```


### Abbreviations

-   **LP**: Laura Perlaza-Jiménez
-   **HK**: Hossein Valipour Kahrood
-   **GI**: Giulia Iacono
-   **SA**: Stuart Archer
