# Enrichment Statistics



Enrichment statistics are based on a contingency table, as shown below:

<!-- | Gene | ..in term | ..not in term | Total | -->
<!-- |:------------------:|----------:|--------------:|------:| -->
<!-- | ..in gene list | 30 | 270 |   300 | -->
<!-- | ..not in gene list (but in background) | 470 | 15230 | 15700 | -->
<!-- | Total |       500 | 15500 | 16000 | -->


<table border="1" style="border-collapse: collapse; width: 100%;">
  <tr>
    <th></th>
    <th>..in term</th>
    <th>..not in term</th>
    <td>Total</td>
  </tr>
  <tr>
    <th>..in gene list</th>
    <td style="background-color: #ffcccb;">30</td>
    <td style="background-color: #90ee90;">270</td>
    <td>300</td>
  </tr>
  <tr>
    <th>..not in gene list (but in background)</th>
    <td style="background-color: #add8e6;">470</td>
    <td style="background-color: #ffffe0;">15230</td>
    <td>15700</td>
  </tr>
  <tr>
    <td>Total</td>
    <td>500</td>
    <td>15500</td>
    <!-- <td>16000</td> -->
    <td style="color: orange;">16000</td> <!-- Colored text for 16000 -->
  </tr>
</table>


This is based on the 16,000 genes that were measured in your experiment. 

Note that some genes may not have been measured; these are excluded entirely from the calculations.
For example, there might have been an additional 5,000 genes (some of which could have been annotated with the term of interest), resulting in a total of *21,250 annotated genes*.

---

## <span style="color:orange;">Fisher's Exact Test</span>

Fisher’s Exact Test is a statistical test used to determine whether there are non-random associations between the proportions of two categorical variables. It calculates the exact probability of observing the given distribution of counts in a 2x2 contingency table, under the null hypothesis of no association between the variables.


> _Note:_ This is just a toy calculator for this training and is quite limited. You can also use some online tools like [Social Science Statistics](https://www.socscistatistics.com/tests/fisher/default2.aspx) to play with.

*Formula*:
$$P = \frac{(a + b)!(c + d)!(a + c)!(b + d)!}{a!b!c!d!N!}$$

Where:

- \(a\), \(b\), \(c\), and \(d\) are the observed counts in the 2x2 contingency table.

- \(N\) is the total number of observations, \(N = a + b + c + d\).

Given this contingency table:

|            | Category 1 | Category 2 | Total |
|------------|------------|------------|-------|
| **Group 1**   | \(a\)        | \(b\)        | \(a + b\) |
| **Group 2**   | \(c\)        | \(d\)        | \(c + d\) |
| **Total**     | \(a + c\)    | \(b + d\)    | \(a + b + c + d\) |


<details>
<summary><em>R syntax</em></summary>

```{r eval=TRUE, echo=TRUE, results='hide', message=FALSE}
a = 30
b = 270
c = 470
d = 15230

data <- matrix(c(a, b, c, d), nrow = 2, byrow = TRUE)

# One-tailed test
fisher.test(data, alternative = "g") # greater or "less"
```
</details> 


## <span style="color:orange;">Hypergeometric Test</span>


The hypergeometric test calculates the probability of observing the given number of genes from a specific category (e.g., a pathway) in the gene list (differentially expressed genes) by chance; it models the situation where you draw a sample (the gene list) from a finite population (the background of all genes), and success is defined as a gene being in the category (e.g., belonging to the pathway).


> _Note:_ Here is a tool by [Stat Trek](https://stattrek.com/online-calculator/hypergeometric) to experiment with the hypergeometric test.

*Formula:*
$$P(X = k) = \frac{\binom{K}{k} \binom{N - K}{n - k}}{\binom{N}{n}}$$
Where:

  - \(k\) = Number of success items in the sample (overlap).
  
  - \(n\) = Number of items in the sample (DE genes).
  
  - \(K\) = Number of success items in the population (in term).
  
  - \(N\) = Total number of items in the population (background).


The parameters in our example:
k=30; n=300; K=500; N=16000;

Where:

`k−1` is the number of observed successes minus 1 (for the "at least" scenario).
Setting `lower.tail = FALSE` gives the probability of obtaining at least 𝑘 successes (right-tail).


<details>
<summary><em>R syntax</em></summary>

```{r eval=TRUE, echo=TRUE, results='hide', message=FALSE}
# Observed value
k <- a
n <- a + b
K <- a + c
N <- a + b + c + d

# (P[X >= k])
phyper(k - 1, K, N - K, n, lower.tail = FALSE)
```
</details>

---

####  {-}

## Activity

### **Challenge:** Interactive Calculator {- .challenge}

[*Link to open toy enrichment calculator*](https://bioinformatics3.erc.monash.edu/rsconnect/content/241/). 

This calculates enrichment for a single hypothetical gene list (e.g. your RNA-seq differentially expressed gene-list) against a single hypothetical 'term' (a set of interesting genes, e.g., synaptic signalling genes). It makes a Venn diagram and a wordy description of what is being tested.

You can adjust various factors and see their effect on the enrichment p-values. 

####  {-}

### **Questions** {- .rationale}
1. If 24 of the 300 differentially expressed genes are annotated with the 500-gene term of interest. Is it significant at p=0.05?
<p>&nbsp;</p>  <!-- insert blank line -->

2. What about with a smaller background of 5000 genes (e.g. proteomic datasets)?
<p>&nbsp;</p>  <!-- insert blank line -->

3. Or, testing against a smaller database of terms; 2000 terms instead of 10000? With the original 16000 gene background.
<p>&nbsp;</p>  <!-- insert blank line -->

4. 19 out of 200 differentially expressed genes (9.5%), need to hit for a 500-gene term (3.1% of all genes) to be significant at (p=0.048). How many hits would be needed for a more specific 30-gene term?

####  {-}

## <span style="color:orange;">Kolmogorov–Smirnov (KS)–like Test</span>
The Kolmogorov–Smirnov (KS)–like test is used in Functional Class Scoring approaches such as Gene Set Enrichment Analysis (GSEA) to test whether genes from a predefined set are non-randomly distributed near the top or bottom of a ranked list of all genes.

This test computes a running-sum statistic as it walks down the ranked list:

 - The running sum increases when a gene in the set is encountered (weighted by the strength of its rank statistic).

 - It decreases when a gene not in the set is encountered.
 
 - The maximum deviation from zero is the enrichment score (ES).
 
The significance of the observed ES is then estimated by *permutation testing*, generating a null distribution to calculate p-values, followed by FDR correction for multiple comparisons.


<details>
<summary>*Formula:*</summary>

According to [Subramanian et al. (2005)](https://www.pnas.org/doi/full/10.1073/pnas.0506580102), the enrichment score is defined as:

$$
P_{\text{hit}}(S,i) = \sum_{\substack{g_j \in S \\ j \le i}} \frac{|r_j|^{p}}{N_R}
$$

$$
P_{\text{miss}}(S,i) = \sum_{\substack{g_j \notin S \\ j \le i}} \frac{1}{N - N_H}
$$

$$
\mathrm{ES}(S) = \max_i \big[ P_{\text{hit}}(i) - P_{\text{miss}}(i) \big]
$$

Explanation of the terms:

| Symbol                               | Meaning                                                   |
|:-------------------------------------|:----------------------------------------------------------|
| \( S \)                              | The gene set being tested (e.g., pathway or GO term). |
| \( L = \{ g_1, g_2, \dots, g_N \} \) | The ranked list of all genes, ordered by correlation with the phenotype. |
| \( g_j \)                            | The j-th gene in the ranked list. |
| \( r_j \)                            | The ranking metric for gene \( g_j \) (e.g., signal-to-noise ratio, t-statistic, or correlation). |
| \( p \)                              | The weighting parameter (usually \( p = 1 \) in standard GSEA). Higher \( p \) values emphasize genes with stronger correlations. |
| \( N_H = |S| \)                      | The number of genes in the gene set \( S \). |
| \( N \)                              | The total number of genes in the ranked list. |
| \( N_R = \sum_{g_j \in S} |r_j|^p \) | The normalisation constant for the “hits” in the gene set. |
| \( P_{\text{hit}}(i) \)              | The cumulative fraction of genes in \( S \) found up to position \( i \). |
| \( P_{\text{miss}}(i) \)             | The cumulative fraction of genes not in \( S \) found up to position \( i \). |
| \( \mathrm{ES}(S) \)                 | The enrichment score, i.e., the maximum deviation between \( P_{\text{hit}}(i) \) and \( P_{\text{miss}}(i) \). A positive ES indicates enrichment at the top of the ranked list; a negative ES indicates enrichment at the bottom. |

*(Adapted from Subramanian et al., 2005, Proc. Natl. Acad. Sci. U.S.A. 102:15545–15550.)*

</details>



