# FGSEA

```{r fgsea-echo-setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

[fgsea](https://bioconductor.org/packages/release/bioc/html/fgsea.html) ("fast gene set enrichment analysis") is an R/Bioconductor package for running pre-ranked GSEA. Rather than testing a fixed list of "interesting" genes as ORA does, GSEA uses the ranks (or ranking statistic, e.g. log2 fold change or t-statistic) of **all** measured genes to ask whether a gene set is skewed towards the top or bottom of that ranking.

`fgsea` implements the same underlying pre-ranked GSEA method as the `GSEA Desktop`/`MSigDB` web tool you used earlier in this workshop, but as a fast, scriptable R function, using an adaptive multi-level Monte-Carlo scheme that lets it accurately estimate very small p-values without needing a large number of permutations. See the [fgsea preprint](https://www.biorxiv.org/content/10.1101/060012v3) for algorithmic detail, and the [package vignette](https://bioconductor.org/packages/release/bioc/vignettes/fgsea/inst/doc/fgsea-tutorial.html) for the full reference this chapter is based on.

We will re-run pre-ranked GSEA against the MSigDB Hallmark gene sets, using the **same pre-ranked gene list** you used earlier in the GSEA Desktop tool (`Pre_Ranked_List_logFC_SYMBOL`), so we can compare the R output directly against what you saw on the web tool, including the `HALLMARK_CHOLESTEROL_HOMEOSTASIS` gene set discussed in the earlier challenge question.

<p> </p>

**In this chapter we will:**

1.  Load the pre-ranked gene list used in the earlier GSEA Desktop activity
2.  Prepare the ranked statistics vector required by `fgsea`
3.  Retrieve the MSigDB Hallmark gene sets with `msigdbr`
4.  Run pre-ranked GSEA with the `fgsea` function
5.  Save the tabular results to a file
6.  Visualise an individual gene set with `plotEnrichment`
7.  Visualise many gene sets at once with `plotGseaTable`
8.  Collapse redundant/overlapping gene sets with `collapsePathways`
9.  Run an over-representation test with `fora` for comparison against ORA

All of the data used in this chapter is provided in the `data/` folder of this workshop's materials, so there's nothing extra to download - just read on and run the code chunks below in order (in RStudio, open this same `.Rmd` file directly to run and edit the code yourself).

<p> </p>

## Load R packages

```{r fgsea-packages, message=FALSE}
library(readr)
library(dplyr)
library(fgsea)
library(data.table)
library(ggplot2)
library(msigdbr)
library(BiocParallel)

register(SerialParam())
```

## 1. Load input data

Earlier, we used a pre-ranked gene list derived from the [Pezzini et al. 2017](https://link.springer.com/article/10.1007/s10571-016-0403-y) SH-SY5Y neuronal differentiation dataset: genes ranked by their log2 fold change between differentiated and undifferentiated cells (`Pre_Ranked_List_logFC_SYMBOL`).

We load that same file here so that our `fgsea` results are directly comparable to what you saw on the web tools.

```{r fgsea-load-pre-ranked}
ranks_df <- read_csv("data/STRING/Pre_Ranked_List_logFC_SYMBOL", show_col_types = FALSE)
head(ranks_df)
```

The file has one row per gene, with `Gene_Symbol` and `logFC` columns, already sorted from most strongly up-regulated to most strongly down-regulated in the differentiated condition.

## 2. Prepare the ranked statistics vector

`fgsea` requires the ranking statistics as a single **named numeric vector** (gene symbol -\> statistic), rather than a data frame. A small number of gene symbols in the table are duplicated (for example where multiple probes/transcripts map to the same gene symbol); for these we keep the value with the largest absolute log fold change.

```{r fgsea-build-ranks}
ranks_df <- ranks_df %>%
  group_by(Gene_Symbol) %>%
  slice_max(abs(logFC), n = 1, with_ties = FALSE) %>%
  ungroup()

ranks <- setNames(ranks_df$logFC, ranks_df$Gene_Symbol)

# fgsea expects the vector sorted, largest to smallest
ranks <- sort(ranks, decreasing = TRUE)

str(ranks)
```

## 3. Get MSigDB Hallmark gene sets

Earlier, the GSEA Desktop and STRING GSEA activities queried the [MSigDB](https://www.gsea-msigdb.org/gsea/msigdb) Hallmark gene set collection. The [`msigdbr`](https://cran.r-project.org/web/packages/msigdbr/index.html) package provides convenient, versioned, tidy access to MSigDB collections directly in R, without needing to download `.gmt` files by hand.

```{r fgsea-get-hallmark}
hallmark_df <- msigdbr(species = "Homo sapiens", category = "H")
head(hallmark_df)
```

`fgsea` expects pathways as a **named list** of character vectors (gene set name -\> genes), so we reshape the `msigdbr` table into that format:

```{r fgsea-reshape-pathways}
pathways <- split(hallmark_df$gene_symbol, hallmark_df$gs_name)
str(head(pathways, 2))
```

## 4. Run fgsea

Before running the below code chunk, review the parameters for the `fgsea` function using the `Help` pane of RStudio (`?fgsea::fgsea`). Note the similarities to the parameters you saw in the GSEA Desktop tool, e.g. `minSize`/`maxSize` (gene set size filters).

```{r fgsea-help, eval=FALSE}
?fgsea::fgsea
```

```{r fgsea-run}
set.seed(42)

fgseaRes <- fgsea(pathways = pathways,
                   stats    = ranks,
                   minSize  = 15,
                   maxSize  = 500)
```

`fgsea` has a default lower bound (`eps = 1e-10`) on the p-values it estimates, for speed. If a gene set could plausibly have an even smaller p-value than this, you will see a warning. To estimate p-values more precisely (at the cost of a slower run), set `eps = 0`:

```{r fgsea-run-eps0}
fgseaRes <- fgsea(pathways = pathways,
                   stats    = ranks,
                   eps      = 0,
                   minSize  = 15,
                   maxSize  = 500)

head(fgseaRes[order(pval), ])
```

The result table contains, per gene set: the enrichment score (`ES`), normalised enrichment score (`NES`), raw and adjusted p-values, and the `leadingEdge` genes driving the enrichment.

Let's check whether `HALLMARK_CHOLESTEROL_HOMEOSTASIS` (discussed in the earlier challenge question) shows up as significantly enriched here too:

```{r fgsea-check-cholesterol}
fgseaRes[pathway == "HALLMARK_CHOLESTEROL_HOMEOSTASIS"]
```

## 5. Save the results to a file

`data.table::fwrite` handles the list-column (`leadingEdge`) automatically, unlike `write.csv`/`write.table`.

```{r fgsea-write-results}
title <- "fgsea_Hallmark_results.txt"
fwrite(fgseaRes, file = title, sep = "\t", sep2 = c("", " ", ""))
cat("Table saved to", title, "\n")
```

## 6. Visualise a single gene set

`plotEnrichment` reproduces the classic GSEA "running enrichment score" plot for one gene set, the same style of plot produced by the GSEA Desktop tool:

```{r fgsea-plot-enrichment-cholesterol, fig.width=7, fig.height=4}
plotEnrichment(pathways[["HALLMARK_CHOLESTEROL_HOMEOSTASIS"]], ranks) +
  labs(title = "HALLMARK_CHOLESTEROL_HOMEOSTASIS")
```

## 7. Visualise the top gene sets as a table plot

`plotGseaTable` draws a compact summary of many gene sets at once, showing each gene set's ranked-list position, running score trace, and NES/p-value.

```{r fgsea-select-top-pathways}
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n = 10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n = 10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
```

```{r fgsea-plot-gsea-table, fig.width=10, fig.height=8, fig.retina=2}
plotGseaTable(pathways[topPathways], ranks, fgseaRes, gseaParam = 0.5)
```

## 8. Collapse redundant gene sets

Hallmark gene sets are curated to minimise overlap, but many gene set collections (e.g. GO, Reactome) contain highly overlapping terms. `collapsePathways` identifies a smaller, largely independent subset of the significant gene sets whose leading-edge genes are not already explained by another, more significant gene set:

```{r fgsea-collapse-pathways, fig.width=10, fig.height=8, fig.retina=2, warning=FALSE}
collapsedPathways <- collapsePathways(fgseaRes[order(pval)][padj < 0.05],
                                       pathways, ranks)
mainPathways <- fgseaRes[pathway %in% collapsedPathways$mainPathways][
  order(-NES), pathway]

plotGseaTable(pathways[mainPathways], ranks, fgseaRes, gseaParam = 0.5)
```

## 9. Over-representation test with fora

`fgsea` also includes `fora`, a fast implementation of a classic hypergeometric over-representation test (like the ORA you ran earlier with `g:Profiler` and `STRING`). It requires a foreground gene list, a background/universe gene list, and a set of pathways.

We reuse the same `DE_198set_Gene_Symbols.txt` gene list you used for ORA in `g:Profiler` and `STRING` earlier as our foreground, and all genes measured in the experiment as our background, so we can compare the GSEA and ORA results for this dataset directly.

```{r fgsea-fora}
fg <- read_lines("data/STRING/DE_198set_Gene_Symbols.txt")
bg <- names(ranks)

foraRes <- fora(genes = fg, universe = bg, pathways = pathways)
head(foraRes[order(pval), ])
```

<p style="color:#006400;">

**Question:** Do the top gene sets from `fora` (ORA) match the top gene sets from `fgsea` (GSEA)? Why might they differ, given that ORA only sees the 198 significant genes while GSEA uses the ranking of all \~14,000 genes?

</p>

## Performance considerations

`fgsea` is parallelised via the `BiocParallel` package. By default, the first registered backend returned by `bpparam()` is used (above we explicitly registered `SerialParam()` for reproducibility). To use multiple cores, set the `nproc` argument of `fgsea()`, or supply your own `BPPARAM`, e.g. `BPPARAM = MulticoreParam(workers = 4)`.

<p> </p>
