# FGSEA

```{r fgsea-echo-setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

[fgsea](https://bioconductor.org/packages/release/bioc/html/fgsea.html) ("fast gene set enrichment analysis") is an R package for running pre-ranked GSEA. GSEA uses the ranks (or ranking statistic, e.g. log2 fold change or t-statistic) of **all** measured genes to ask whether a gene set is skewed towards the top or bottom of that ranking.

`fgsea` implements the same underlying pre-ranked GSEA method as the `GSEA Desktop`/`MSigDB` web tool you used in the previous chapter, but as a fast, scriptable R function. See the [fgsea preprint](https://www.biorxiv.org/content/10.1101/060012v3) for algorithmic detail, and the [package vignette](https://bioconductor.org/packages/release/bioc/vignettes/fgsea/inst/doc/fgsea-tutorial.html) for the full reference this chapter is based on.

## Input data

We will re-run pre-ranked GSEA against the MSigDB Hallmark gene sets, using the same pre-ranked gene list you used earlier (`Pre_Ranked_List_logFC_SYMBOL`), so we can compare the R output directly against what you saw in the previous chapter, including the `HALLMARK_CHOLESTEROL_HOMEOSTASIS` gene set discussed in the earlier challenge question.  
All of the data used in this chapter is provided in the `data/` folder of this workshop's materials, so there's nothing extra to download - just read on and run the code chunks below in order.

<p> </p>

## Activity overview

::: {.overview}
1.  Load the pre-ranked gene list 
2.  Prepare the ranked statistics vector required by `fgsea`
3.  Retrieve the MSigDB Hallmark gene sets with `msigdbr`
4.  Run pre-ranked GSEA with the `fgsea` function
5.  Save the tabular results to a file
6.  Visualise an individual gene set with `plotEnrichment`
7.  Visualise many gene sets at once with `plotGseaTable`
8.  Collapse redundant/overlapping gene sets with `collapsePathways`
:::

## Load R packages

```{r fgsea-packages, message=FALSE}
library(readr)
library(dplyr)
library(fgsea)
library(data.table)
library(ggplot2)
library(msigdbr)
library(BiocParallel)

register(SerialParam())
```

## Load input data

We load the DE genes table derived from the [Pezzini et al. 2017](https://link.springer.com/article/10.1007/s10571-016-0403-y) SH-SY5Y neuronal differentiation dataset. The file has one row per gene, with `Gene_Symbol` and `logFC` columns and other related statistics columns, comparing differentiated against undifferentiated cells (`Pre_Ranked_List_logFC_SYMBOL`).

```{r fgsea-load-pre-ranked}
stats_df <- read.table("day2_inputData/Pezzini_DE.txt", header = TRUE)
head(stats_df)
```

## Prepare the ranked statistics vector

`fgsea` requires the ranking statistics as a single **named numeric vector** (gene id -\> statistic), so we extract the log2 fold change column as a named vector, and arrange it by decreasing log2 fold change. Here, genes at the start of this vector reflect a strong increase in the differentiated cells compared to the undifferentiated cells, while genes at the bottom of this list are strongly decreased in the differentiated cells.

```{r fgsea-build-ranks}
stats <- setNames(stats_df$Log2FC, stats_df$Gene.ID)

# fgsea expects the vector sorted, largest log2 fold change to smallest
stats <- sort(stats, decreasing = TRUE)
```

## Get MSigDB Hallmark gene sets

Earlier, we queried the [MSigDB](https://www.gsea-msigdb.org/gsea/msigdb) Hallmark gene set collection. The [`msigdbr`](https://cran.r-project.org/web/packages/msigdbr/index.html) package provides convenient, versioned, tidy access to MSigDB collections directly in R, without needing to download `.gmt` files by hand. There are 50 unique Hallmark gene sets in this collection. After running the code below, have a look at what they are by running `unique(hallmark_df$gs_name)` in your console.

```{r fgsea-get-hallmark}
hallmark_df <- msigdbr(species = "Homo sapiens", collection = "H")
head(hallmark_df)
```

`fgsea` expects pathways as a **named list** of character vectors (gene set name -\> genes), so we reshape the `msigdbr` table into that format. Here we use the ensembl gene IDs, as some of the gene symbols in the input table are duplicated.

```{r fgsea-reshape-pathways}
pathways <- split(hallmark_df$ensembl_gene, hallmark_df$gs_name)
str(head(pathways))
```

## Run fgsea

Before running the below code chunk, review the parameters for the `fgsea` function using the `Help` pane of RStudio (`?fgsea::fgsea`). Note the similarities to the parameters you saw in the GSEA Desktop tool, e.g. `minSize`/`maxSize` (gene set size filters).

```{r fgsea-help, eval=FALSE}
?fgsea::fgsea
```

Let's run fgsea:

```{r fgsea-run}
set.seed(42)

fgseaRes <- fgsea(pathways = pathways,
                  stats    = stats, #	Named vector of gene-level stats. Names should be the same as in 'pathways'
                  minSize  = 15, # 	Minimal size of a gene set to test. All pathways below the threshold are excluded.
                  maxSize  = 500, #	Maximal size of a gene set to test. All pathways above the threshold are excluded.
                  eps = 1e-10)

#`fgsea` has a default lower bound (`eps = 1e-10`) on the p-values it estimates, for speed. If a gene set could plausibly have an even smaller p-value than this, you will see a warning. To estimate p-values more precisely (at the cost of a slower run), set `eps = 0`
```
The result table contains, per gene set: the enrichment score (`ES`), normalised enrichment score (`NES`), raw and adjusted p-values, and the `leadingEdge` genes driving the enrichment.

```{r fgsea-run}
head(fgseaRes[order(pval), ])
```

Let's check whether `HALLMARK_CHOLESTEROL_HOMEOSTASIS` (discussed in the earlier challenge question) shows up as significantly enriched here too:

```{r fgsea-check-cholesterol}
fgseaRes[pathway == "HALLMARK_CHOLESTEROL_HOMEOSTASIS"]
```

## Save the results to a file

`data.table::fwrite` handles the list-column (`leadingEdge`) automatically, unlike `write.csv`/`write.table`.

```{r fgsea-write-results, eval=FALSE}
title <- "fgsea_Hallmark_results.txt"
fwrite(fgseaRes, file = title, sep = "\t", sep2 = c("", " ", ""))
cat("Table saved to", title, "\n")
```

## Visualise a single gene set

`plotEnrichment` reproduces the classic GSEA "running enrichment score" plot for one gene set, the same style of plot produced in the previous chapter:

```{r fgsea-plot-enrichment-cholesterol, fig.width=7, fig.height=4}
plotEnrichment(pathways[["HALLMARK_CHOLESTEROL_HOMEOSTASIS"]], stats) +
  labs(title = "HALLMARK_CHOLESTEROL_HOMEOSTASIS")
```

## Visualise the top gene sets as a table plot

`plotGseaTable` draws a compact summary of many gene sets at once, showing each gene set's ranked-list position, running score trace, and NES/p-value.

```{r fgsea-plot-gsea-table, fig.width=10, fig.height=8, fig.retina=2}
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n = 10), pathway] # filter table by ES > 0, order by increasing pvalue and select top 10 pathways
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n = 10), pathway] # filter table by ES < 0, order by increasing pvalue and select top 10 pathways
topPathways <- c(topPathwaysUp, rev(topPathwaysDown)) # arrange from highest to lowest

plotGseaTable(pathways[topPathways], stats, fgseaRes, gseaParam = 0.5)
```

## Collapse redundant gene sets

Hallmark gene sets are curated to minimise overlap, but many gene set collections (e.g. GO, Reactome) contain highly overlapping terms. `collapsePathways` identifies a smaller, largely independent subset of the significant gene sets whose leading-edge genes are not already explained by another, more significant gene set:

```{r fgsea-collapse-pathways, fig.width=10, fig.height=8, fig.retina=2, warning=FALSE}
collapsedPathways <- collapsePathways(fgseaRes[order(pval)][padj < 0.05], pathways, stats)
mainPathways <- fgseaRes[pathway %in% collapsedPathways$mainPathways][order(-NES), pathway]

plotGseaTable(pathways[mainPathways], stats, fgseaRes, gseaParam = 0.5)
```

## Performance considerations

`fgsea` is parallelised via the `BiocParallel` package. By default, the first registered backend returned by `bpparam()` is used (above we explicitly registered `SerialParam()` for reproducibility). To use multiple cores, set the `nproc` argument of `fgsea()`, or supply your own `BPPARAM`, e.g. `BPPARAM = MulticoreParam(workers = 4)`.

