# STRINGdb

```{r stringdb-echo-setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

Earlier in the book you used the [STRING website](https://string-db.org) directly to explore both a protein-protein interaction network and functional enrichment for a gene list. The [`STRINGdb`](https://bioconductor.org/packages/release/bioc/html/STRINGdb.html) R package gives scripted, reproducible access to the same underlying database and, being an R package, integrates directly with your existing gene lists, expression values, and enrichment workflows. STRINGdb uses R's [Reference Classes](https://stat.ethz.ch/R-manual/R-devel/library/methods/html/refClasses.html) system, so most functionality is accessed as *methods* on a `STRINGdb` object (`string_db$method_name(...)`) rather than as standalone functions, and represents the network internally using [igraph](https://igraph.org/r/).

This chapter is based on the [STRINGdb package vignette](https://bioconductor.org/packages/release/bioc/vignettes/STRINGdb/inst/doc/STRINGdb.pdf), reworked to use our own [Pezzini et al. 2017](https://link.springer.com/article/10.1007/s10571-016-0403-y) SH-SY5Y neuronal differentiation dataset. Note that several `STRINGdb` methods (network images, enrichment figures, the payload colouring) query the STRING web server directly, so an internet connection is required.

<p> </p>

## Activity overview

::: {.overview}
1.  Initialise a `STRINGdb` object for human (NCBI taxonomy ID 9606)
2.  Load the differential expression results and map gene symbols to STRING identifiers
3.  Visualise the protein-protein interaction network
4.  Colour the network by up/down-regulation with the payload mechanism
5.  Run functional enrichment across GO, KEGG, Reactome and other databases
6.  Set a custom statistical background
7.  Retrieve functional annotations for the gene set
8.  Cluster the network and visualise individual clusters
9.  Query specific proteins of interest: interactions, interaction partners, paralogs and homologs
:::

## Load R packages

```{r stringdb-packages, message=FALSE}
library(readr)
library(dplyr)
library(STRINGdb)
```

## Initialise the STRINGdb object

To begin, you should know the NCBI taxonomy identifier of your organism of interest (e.g. `9606` for human, `10090` for mouse). If you don't know it, search [NCBI Taxonomy](https://www.ncbi.nlm.nih.gov/taxonomy) or the [STRING organisms list](https://string-db.org/cgi/input.pl?input_page_active_form=organisms).

In the constructor below we also set:

- `score_threshold`: the minimum combined confidence score (0-1000) for an interaction to be loaded (400 is STRING's own "medium confidence" default, and matches what you used on the website)
- `network_type`: `"full"` for the full functional network (as opposed to `"physical"`, which restricts to physical complexes only)
- `link_data`: `"detailed"`, so that individual evidence scores (e.g. experimental, database, text-mining) are loaded alongside the combined score

```{r stringdb-initialise}
string_db <- STRINGdb$new(version = "12.0", species = 9606,
                           score_threshold = 400, network_type = "full",
                           link_data = "detailed", input_directory = "")
```

## Explore the available methods

Since `STRINGdb` uses Reference Classes, all its functionality is exposed as methods on the `string_db` object. Two commands are useful for exploring what's available:

```{r stringdb-explore-methods, eval=FALSE}
STRINGdb$methods()              # list all available methods
STRINGdb$help("get_graph")      # view documentation for a specific method
```

For every method used below, you can call `STRINGdb$help("method_name")` for the full list of parameters and defaults.

## Load the differential expression data

We reuse the same differential expression results behind the `DE_198set` gene list you used for ORA in `g:Profiler` and `STRING` earlier: genes from the Pezzini et al. 2017 SH-SY5Y differentiation experiment, each with a log2 fold change and an FDR-adjusted p-value.

```{r stringdb-load-de}
de_full <- read.table("data/R_data/Pezzini_DE.txt", header = TRUE)
head(de_full)
```

The earlier `DE_198set` gene list was obtained by filtering this table for FDR \< 0.0001 and an absolute log2 fold change ≥ 1 (i.e. at least a 2-fold change). We recreate that same filtered set here, keeping the `pvalue` (we'll use FDR as our p-value, as in the STRING web activity) and `logFC` columns, so we can later colour the network by direction of change:

```{r stringdb-filter-de}
de_198 <- de_full %>%
  filter(FDR < 0.0001 & abs(Log2FC) >= 1)

cat("Number of genes passing FDR and fold change filter:", nrow(de_198), "\n")
head(de_198)
```

## Map gene identifiers to STRING IDs

As a first step, we map our gene identifiers to STRING database identifiers using the `map` method. Here we map from HUGO gene symbols, but `map` also supports other common identifiers (e.g. Entrez Gene ID, Ensembl protein, RefSeq transcript, etc).

The `map` method adds a `STRING_id` column to the data frame passed as its first argument.

```{r stringdb-map-genes}
de_198_mapped <- string_db$map(de_198, "Gene.Name", removeUnmappedRows = TRUE)
head(de_198_mapped)
```

If a warning about unmapped genes is printed, this is expected: not every symbol will resolve to a STRING identifier (for example, deprecated gene symbols or non-protein-coding genes). Setting `removeUnmappedRows = FALSE` instead would keep the unmapped rows so you can inspect them manually.

## Visualise the interaction network

We extract the STRING identifiers as our `hits`, and plot the network. The p-value printed above the plot is the probability of observing at least this many interactions among these proteins by chance - the same statistic reported on the STRING website's network page.

```{r stringdb-plot-network, fig.width=8, fig.height=8}
hits <- de_198_mapped$STRING_id

string_db$plot_network(hits)
```

Compare this network to the one you generated on the STRING website earlier using the same `DE_198set` gene list - the nodes, edges, and clustering should look familiar!

If you want more control over the appearance of the network image, `get_png` and `get_link` expose additional styling options, letting you switch between `evidence`/`confidence` edge styles, hide labels or disconnected proteins, use a flat node design, and adjust label size.

## Payload mechanism: colour by direction of change

STRINGdb supports the "payload" mechanism used on the website to add a coloured halo around network nodes - for example, colouring down-regulated genes green and up-regulated genes red. This works by posting the colour information to the STRING server, then requesting the network image again with the resulting `payload_id`.

```{r stringdb-add-diff-exp-colour, fig.width=8, fig.height=8}
de_198_coloured <- string_db$add_diff_exp_color(de_198_mapped, logFcColStr = "Log2FC")

payload_id <- string_db$post_payload(de_198_coloured$STRING_id,
                                      colors = de_198_coloured$color)

string_db$plot_network(hits, payload_id = payload_id)
```

## Functional enrichment

`get_enrichment` computes enrichment across Gene Ontology (Process, Function, Component), KEGG, Reactome, PubMed publications, UniProt Keywords, and PFAM/InterPro/SMART domains, in a single call. It uses a hypergeometric test with Benjamini-Hochberg FDR correction - the same underlying statistics as the STRING website enrichment tab you used earlier.

```{r stringdb-enrichment}
enrichment <- string_db$get_enrichment(hits)
head(enrichment, n = 20)
```

<p style="color:#006400;">

**Question:** How closely do these results match the enrichment table you downloaded from the STRING website earlier for the same gene list?

</p>

You can also retrieve a ready-made enrichment figure directly from STRING for a chosen category:

```{r stringdb-enrichment-figure, fig.width=8, fig.height=6}
enrichment_img <- string_db$get_enrichment_figure(
  hits,
  category = "Component",
  group_by_similarity = 0.8,
  x_axis = "signal"
)

grid::grid.newpage()
grid::grid.raster(enrichment_img, interpolate = FALSE)
```

## Setting a custom statistical background

As discussed earlier in the book, using an appropriate background/universe is important for avoiding false positives in enrichment tests. By default, `get_enrichment` tests against the whole proteome. If your experiment only measured a defined subset of genes (as ours does), you should set that as the background instead:

```{r stringdb-set-background}
background <- string_db$map(transmute(de_full, gene = Gene.Name), "gene", removeUnmappedRows = TRUE)$STRING_id
string_db$set_background(background)
```

Once a background is set, it is used automatically by subsequent calls to `get_enrichment` on the same `string_db` object - you do not need to create a new object.

```{r stringdb-enrichment-with-background}
enrichment_background_corrected <- string_db$get_enrichment(hits)
head(enrichment_background_corrected, n = 20)
```

Compare this to the enrichment table from the previous step: results are often less dramatic once tested against a realistic background rather than the whole proteome, because the test is now calibrated against the genes that were actually measurable in this experiment.

## Retrieve functional annotations

If you simply want to know which terms are *assigned* to your proteins, rather than which are statistically enriched, use `get_annotations`. This returns terms from most categories associated with your protein set (KEGG and PubMed are excluded, for licensing and output-size reasons respectively).

```{r stringdb-annotations}
annotations <- string_db$get_annotations(hits)
head(annotations, n = 20)
```

## Clustering

The underlying `igraph` package provides several clustering/community-detection algorithms (`"fastgreedy"`, `"walktrap"`, `"spinglass"`, `"edge.betweenness"`), which `get_clusters` wraps in an easy-to-use function returning a list of clusters.

```{r stringdb-clustering}
clustersList <- string_db$get_clusters(hits)
```

We visualise the first four clusters individually:

```{r stringdb-plot-clusters, fig.width=6, fig.height=6}
par(mar = c(0, 0, 0.5, 0))
string_db$plot_network(clustersList[[1]], add_summary = FALSE)

par(mar = c(0, 0, 0.5, 0))
string_db$plot_network(clustersList[[2]], add_summary = FALSE)

par(mar = c(0, 0, 0.5, 0))
string_db$plot_network(clustersList[[3]], add_summary = FALSE)

par(mar = c(0, 0, 0.5, 0))
string_db$plot_network(clustersList[[4]], add_summary = FALSE)
```

## Querying specific proteins of interest

`CRABP2` and `SEMA3A` are among the most significantly differentially expressed genes in this dataset (see the DE table above). Let's look at them specifically, using `mp` ("map proteins") - a convenient alternative to `map` for mapping just one or a few gene symbols directly to STRING identifiers:

```{r stringdb-map-specific-proteins}
crabp2 <- string_db$mp("crabp2")
sema3a <- string_db$mp("sema3a")
```

### Interactions between named proteins

`get_interactions` returns a data frame describing the interactions among the queried proteins, with STRING identifiers (`from`/`to`), preferred names (`from_name`/`to_name`), the `combined_score`, and any additional evidence scores available under the current `link_data` setting.

```{r stringdb-interactions}
string_db$get_interactions(c(crabp2, sema3a))
```

An empty data frame here is expected, not an error: `CRABP2` and `SEMA3A` are not annotated as directly interacting in STRING, so there is no row to return. `get_interactions` only reports edges that exist directly between the queried proteins; it will not tell you about shared partners or indirect paths between them.

### Interaction partners

To retrieve the first-shell interaction partners of one or more proteins from the locally loaded network, use `get_interaction_partners`. The output has the same structure as `get_interactions`.

```{r stringdb-interaction-partners}
string_db$get_interaction_partners(c(crabp2, sema3a), limit = 5)[, c("from", "to", "combined_score", "experimental", "from_name", "to_name")]
```

You can restrict to interactions supported by a particular type of evidence - here, only those with experimental support:

```{r stringdb-interaction-partners-experimental}
interaction_partners <- string_db$get_interaction_partners(c(crabp2, sema3a), limit = 5)

interaction_partners_exp <- subset(interaction_partners, experimental > 0,
                                    select = c("from_name", "to_name", "experimental"))
interaction_partners_exp
```

<p> </p>
