# Setting Up R Packages

```{r setup-echo-setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

The command-line chapters of this workshop (`gprofiler2`, `clusterProfiler` and `enrichplot`, Novel species FEA, `FGSEA`, and `STRINGdb`) use a number of R packages. On the day, these are already installed on the workshop-provided computers/VMs, so no action is needed. This page is for anyone who wants to run the code on their **own computer**, either during the workshop or afterwards.

This workshop was built and tested with **R 4.6.1** and **Bioconductor 3.23**. Any recent R release (4.4 or later) with a matching Bioconductor release should work; see the [Bioconductor release/R version table](https://bioconductor.org/about/release-announcements/) if you are installing on an older R version.

<p>&nbsp;</p>  <!-- insert blank line -->

## CRAN packages

Most of these come from the [tidyverse](https://www.tidyverse.org/) plus a handful of enrichment-specific packages:

```{r install-cran, eval = FALSE}
install.packages(c(
  "tidyverse",  # includes dplyr, ggplot2, readr, tidyr, and friends
  "data.table",
  "gprofiler2",
  "ggupset",
  "ggtangle",
  "msigdbr",
  "ontologyIndex",
  "WebGestaltR"
))
```

<p>&nbsp;</p>  <!-- insert blank line -->

## Bioconductor packages

Bioconductor packages are installed via `BiocManager` rather than `install.packages()`:

```{r install-bioc, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c(
  "clusterProfiler",
  "enrichplot",
  "biomaRt",
  "org.Hs.eg.db",
  "fgsea",
  "BiocParallel",
  "STRINGdb"
))
```

`org.Hs.eg.db` is the human genome-wide annotation package; it matches the human [Pezzini et al. 2017](https://link.springer.com/article/10.1007/s10571-016-0403-y) dataset used throughout the workshop.

<p>&nbsp;</p>  <!-- insert blank line -->

## Optional: GitHub-only package

The `clusterProfiler`/`enrichplot` chapter mentions `multienrichjam`, a package that converts ORA results from other tools into `enrichplot`-compatible objects. It is **not required** for this workshop, has a large dependency footprint, and is not installed on the workshop VMs. If you'd like to try it on your own computer:

```{r install-multienrichjam, eval = FALSE}
install.packages("remotes")
remotes::install_github("jmw86069/multienrichjam", dependencies = TRUE)
```

<p>&nbsp;</p>  <!-- insert blank line -->

## Where each package is used

::: {.overview}
| Package             | Source        | Used in                                              |
|---------------------|---------------|-------------------------------------------------------|
| `tidyverse`/`dplyr`/`ggplot2`/`readr`/`tidyr` | CRAN | throughout all Day 2 chapters |
| `data.table`         | CRAN          | FGSEA                                                 |
| `gprofiler2`         | CRAN          | gprofiler2                                            |
| `ggupset`            | CRAN          | clusterProfiler and enrichplot                        |
| `ggtangle`           | CRAN          | clusterProfiler and enrichplot (`cnetplot` labelling)  |
| `msigdbr`            | CRAN          | FGSEA                                                 |
| `ontologyIndex`      | CRAN          | Novel species FEA                                     |
| `WebGestaltR`        | CRAN          | Novel species FEA                                     |
| `clusterProfiler`    | Bioconductor  | clusterProfiler and enrichplot, Novel species FEA      |
| `enrichplot`         | Bioconductor  | clusterProfiler and enrichplot, Novel species FEA      |
| `biomaRt`            | Bioconductor  | clusterProfiler and enrichplot                        |
| `org.Hs.eg.db`       | Bioconductor  | clusterProfiler and enrichplot                        |
| `fgsea`              | Bioconductor  | FGSEA                                                 |
| `BiocParallel`       | Bioconductor  | FGSEA                                                 |
| `STRINGdb`           | Bioconductor  | STRINGdb                                              |
| `multienrichjam`     | GitHub (optional) | clusterProfiler and enrichplot (not run)          |
:::

<p>&nbsp;</p>  <!-- insert blank line -->
